---
title: "ESM 244 Lab 1"
author: "Alex Ehrens"
date: "1/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
```

### Read in SF trees data

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"))
```

### Basic wrangling reminders

Refresh some skills for data wrangling and summary statistics using functions in the `dplyr` package.

Find the top 5 highest observations of trees by legal_status, do some wrangling, make a graph.

```{r}
top_5_status <- sf_trees %>% 
  count(legal_status) %>% # count function does all of what group_by, summarize, n() functions do
# can set multiple group levels in same place (count(legal_status, species))
  drop_na(legal_status) %>% # gets rid of rows all NA values in certain variable
  rename(tree_count = n) %>% # new name on left, old name on right
  relocate(tree_count) %>% # moves tree_count to first column position 
  slice_max(tree_count, n = 5) # only leaves top 5 highest values of tree_count
```

Make a grap of those top 5 observations by legal status.
```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), legal_status, y = tree_count)) +
  geom_col() + 
  labs(x = "Legal Status", y = "Tree Count") + 
  coord_flip() +
  theme_minimal()
```

### A few more data wrangling refresher examples

Only want to keep observations (rows) for Blackwood Acacia trees.

```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% #finds any row where "blackwood acacia" string exists within value (doesn't have to be exact match)
  select(legal_status, date, latitude, longitude)

ggplot(data = blackwood_acacia, aes(x = longitude, y = latitude)) +
  geom_point()
```

### tidyr::separate() and unite() functions

Useful for combining or separating columns

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep = "::")
```


Example: tidyr::unite()

```{r}
sf_trees_unite <- sf_trees %>% 
  unite("id_status", tree_id:legal_status, sep = "_cool!_")
```

### Make some actual maps of blackwood acacia trees in SF

`st_as_sf()` to convert latitude & longitude to spatial coordinates.

```{r}
blackwood_acacia_sp <- blackwood_acacia %>% 
  drop_na(longitude, latitude) %>% 
  st_as_sf(coords = c("longitude", "latitude")) # must put coords vector in this order - long (X) first, lat (Y) second

st_crs(blackwood_acacia_sp) = 4326 # set coordinate reference system

ggplot(data = blackwood_acacia_sp) +
  geom_sf(color = "darkgreen")
```

Read in SF roads shapefile:
```{r}
sf_map <- read_sf(here("data", "sf_map", "tl_2017_06075_roads.shp"))

st_transform(sf_map, 4326) # change coordinate system to match other dataset

ggplot(data = sf_map) +
  geom_sf()
```

Combine blackwood acacia tree observations & SF roads map:

```{r}
ggplot() +
  geom_sf(data = sf_map, size = 0.1, color = "darkgray") +
  geom_sf(data = blackwood_acacia_sp, color = "red", size = 0.5) +
  theme_void()
```

Now an interactive map:
```{r}
tmap_mode("view") # sets map to interactive viewing

tm_shape(blackwood_acacia_sp) +
  tm_dots()
```

